---
title: "Bitcoin Price Forecasting Using Time Series Models"
date: "March 9, 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(quantmod)
library(zoo)
library(xts)
library(tseries)
library(forecast)
library(FinTS)
library(rugarch)
library(vars)
library(MTS)
library(tidyverse)
library(reshape2)
library(bsts)
library(PerformanceAnalytics) 
library(scales)                
library(fredr)
```



# Problem Statement

## Objective

Bitcoin (BTC) is the world's largest cryptocurrency by market capitalization and
is known for extreme price volatility — daily swings of 5–15% are common. 
Accurately forecasting BTC's daily price (or returns) would enable traders and 
portfolio managers to identify entry/exit opportunities and better manage risk.

This project asks: 
**Can the inclusion of macroeconomic indicators, market indices, and volatility dynamics  improve Bitcoin forecasts compared to a univariate time-series baseline?**

## Research Questions

1. Does BTC exhibit ARCH/GARCH effects, indicating volatility clustering? If so,
does a VAR-GARCH model forecast BTC better than BSTS or SARIMA, which do not 
explicitly model conditional volatility?

2. Do macroeconomic and market variables—SPY, QQQ, Fed Funds, the 10-Year 
Treasury yield, CPI, and unemployment—Granger-cause BTC returns? If so, do 
multivariate models such as VAR and BSTS improve forecast accuracy relative to a
univariate SARIMA benchmark?

3. Does a BSTS model with regressors, trend, and seasonality outperform a VAR 
model in forecasting BTC?

4. Does a rolling VAR model with GARCH, which adds both time-varying volatility 
modeling and parameter re-estimation relative to the static VAR, lower forecast
error?

## Scope

- **Target variable:** Daily BTC/USD closing price (log-transformed) and log-returns
- **Forecast horizon:** January 2026 onward (rolling 1-step-ahead for VAR/GARCH; direct horizon for SARIMA/BSTS)
- **Training window:** January 2016 – June 2025
- **Testing window:** June 2025 - January 2026



# Assumptions and Hypotheses

## Modeling Assumptions

1. **Log-returns for GARCH/VAR:** Raw prices are 
non-stationary (unit-root), so we work with log-returns 
`r_t = log(P_t / P_{t-1}) × 100` for volatility and multivariate models.
   
2. **Log-prices for SARIMA/BSTS:** Log-prices preserve the interpretation of the 
series in level terms while helping stabilize variance. In SARIMA, stationarity
is achieved through explicit differencing specified in the model (the d 
parameter), while BSTS models nonstationary behavior through latent components 
such as trend and seasonal states rather than explicit differencing.
   
3. **Stationarity after differencing:** We assume one round of log-differencing 
is sufficient to induce stationarity in all series. This is verified empirically
via ADF and KPSS tests.
   
4. **Data Frequency Alignment:** Lower-frequency series—such as CPI and UNRATE 
(monthly) and DGS10 and DFF (weekdays only)—are aligned to a daily frequency 
using forward-filling, with backward-filling applied when no prior value is 
available. This assumes that the most recent observed value remains in effect 
until a new release is available, which is a standard approach when integrating 
macroeconomic data with higher-frequency financial series.
   
5. **Constant model parameters over the training window:** VAR and GARCH 
parameters are assumed stable. Given the 10-year window includes crypto bull and
bear cycles, this is a simplifying assumption.

## Hypotheses

- **H1 (ARCH effect):** BTC returns exhibit significant ARCH effects due to 
volatility clustering (e.g., news-driven spikes). Therefore, a VAR+GARCH model 
is expected to outperform a standard VAR model by explicitly modeling 
time-varying volatility.
- **H2 (Macro linkage):** At least one macroeconomic or market variable 
Granger-causes BTC returns, suggesting that BTC is influenced by broader 
financial conditions. Therefore, multivariate models that incorporate these 
variables, such as VAR and BSTS, are expected to outperform a univariate SARIMA 
model.
- **H3 (Static VAR vs Rolling VAR+GARCH):** A rolling VAR+GARCH forecast, which 
adds both time-varying volatility modeling and parameter re-estimation relative
to the static VAR, will produce lower forecast error.
- **H4 (BSTS flexibility):** Because BSTS jointly models latent trend, 
seasonality, and regression effects, it is expected to provide competitive 
forecasts of BTC log-prices relative to the other approaches.


# Data

## Data Sources

```{r data-fetch}
# Entire Dataset Start and End Date
start_date <- as.Date("2016-01-01")
end_date <- as.Date("2026-01-01")

# Test Set Start Date
test_start <- as.Date("2025-06-01")

# ── Set your FRED API key ──────────────────────────────────────────────────────
# Get a free key at: https://fred.stlouisfed.org/docs/api/api_key.html
# Then replace the string below, or set it via:
#   usethis::edit_r_environ()  and add  FRED_API_KEY=your_key_here
fredr_set_key("32232ec8ba8f84fd7d9e8c535a0f04c9")

# ── Helper: fredr tibble → xts ────────────────────────────────────────────────
fredr_to_xts <- function(series_id, col_name, start = start_date, end = end_date) {
  df <- fredr(series_id     = series_id,
              observation_start = start,
              observation_end = end,
              frequency     = NULL)   # keep native frequency
  x <- xts(df$value, order.by = as.Date(df$date))
  colnames(x) <- col_name
  x
}

# ── Crypto & Equity (quantmod / Yahoo — unchanged) ───────────────────────────
getSymbols("BTC-USD", src = "yahoo", from = start_date, to = end_date)
btc <- Ad(`BTC-USD`); colnames(btc) <- "BTC"

getSymbols("SPY", src = "yahoo", from = start_date, to = end_date)
spy <- Ad(SPY); colnames(spy) <- "SPY"

getSymbols("QQQ", src = "yahoo", from = start_date, to = end_date)
qqq <- Ad(QQQ); colnames(qqq) <- "QQQ"

# ── Macro (fredr) ─────────────────────────────────────────────────────────────
fed    <- fredr_to_xts("DFF",      "FedFunds")   # Fed Funds Rate (daily)
dgs10  <- fredr_to_xts("DGS10",   "DGS10")       # 10-yr Treasury (daily weekdays)
cpi    <- fredr_to_xts("CPIAUCSL","CPI")          # CPI (monthly)
unrate <- fredr_to_xts("UNRATE",  "Unemployment") # Unemployment rate (monthly)
```

**Variable descriptions:**

| Variable       | Source       | Frequency        | Description |
|----------------|--------------|------------------|-------------|
| BTC-USD        | Yahoo        | Daily            | Bitcoin adjusted closing price |
| SPY            | Yahoo        | Daily            | S&P 500 ETF (broad US equity) |
| QQQ            | Yahoo        | Daily            | Nasdaq-100 ETF (tech equity) |
| DFF            | FRED (fredr) | Daily            | Effective Federal Funds Rate |
| DGS10          | FRED (fredr) | Daily (weekdays) | 10-Year Treasury Constant Maturity |
| CPIAUCSL       | FRED (fredr) | Monthly          | Consumer Price Index (all urban consumers) |
| UNRATE         | FRED (fredr) | Monthly          | US Civilian Unemployment Rate |

## Data Merging and Imputation

```{r data-merge}
# Merge all series
daily_data <- merge(btc, spy, qqq, fed, dgs10, cpi, unrate)

# Restrict to start date
daily_data <- daily_data[index(daily_data) >= start_date]

# Expand to full calendar (including weekends / holidays)
full_dates <- seq(from = min(index(daily_data)),
                  to   = max(index(daily_data)),
                  by   = "day")

daily_data <- merge(daily_data, xts(order.by = full_dates), fill = NA)

# Forward fill (carry last known value forward)
daily_data <- na.locf(daily_data, na.rm = FALSE)

# Backward fill for any leading NAs at the start
daily_data <- na.locf(daily_data, fromLast = TRUE)

cat("Dataset dimensions:", dim(daily_data), "\n")
cat("Date range:", as.character(start(daily_data)), "to", as.character(end(daily_data)), "\n")
```

**Imputation rationale:** Markets for BTC trade 24/7 but SPY/QQQ do not trade on 
weekends and US holidays. Macro series (CPI, UNRATE) are released monthly. 
Forward-fill is used because the last observed value is the best available 
estimate until a new release or trading day. No linear interpolation is used to 
avoid look-ahead bias.


# Data Properties and Exploratory Data Analysis (EDA)

```{r eda-function}
# EDA function: summary stats + level plot + returns plot + QQ + ACF/PACF + stationarity tests
eda <- function(x, name, arch_lags = 12) {
  cat("===== EDA:", name, "=====\n")

  x_num <- as.numeric(x)
  x_num <- x_num[is.finite(x_num)]

  print(summary(x_num))
  cat("Mean:     ", round(mean(x_num, na.rm = TRUE), 4), "\n")
  cat("SD:       ", round(sd(x_num, na.rm = TRUE), 4), "\n")
  cat("Skewness: ", round(PerformanceAnalytics::skewness(x_num, na.rm = TRUE), 4),
      " (>0 = right tail, <0 = left tail)\n")
  cat("Kurtosis: ", round(PerformanceAnalytics::kurtosis(x_num, na.rm = TRUE), 4),
      " (>3 = heavy tails relative to normal)\n")

  old_par <- par(no.readonly = TRUE)
  on.exit(par(old_par))
  par(mfrow = c(2, 2))

  # Level plot
  plot.ts(x_num,
          main = paste(name, "— Time Series"),
          col = "steelblue", lwd = 1.2,
          xlab = "Time", ylab = name)

  # QQ plot
  qqnorm(x_num, main = paste("QQ Plot:", name))
  qqline(x_num, col = "red", lwd = 2)

  # ACF/PACF
  Acf(x_num, main = paste("ACF —", name), lag.max = 40)
  Pacf(x_num, main = paste("PACF —", name), lag.max = 40)

  par(mfrow = c(1, 1))

  cat("\nADF Test (H0: unit root):\n")
  print(adf.test(x_num))

  cat("\nKPSS Test (H0: stationary):\n")
  print(kpss.test(x_num))
  
  cat("\nARCH LM Test (H0: no ARCH effects):\n")
  print(FinTS::ArchTest(x_num, lags = arch_lags))
  cat("\n")
}
```

## Target Variable: BTC Log-Returns and Log-Price

```{r log-btc-returns}
# Compute log-returns (%)
btc_ret <- diff(log(daily_data$BTC)) * 100
btc_ret <- na.omit(btc_ret)
```

```{r log-btc-price}
# Compute log-price
btc_log <- log(daily_data$BTC)
btc_log <- na.omit(btc_log)
```

```{r eda-btc-log-returns}
eda(btc_ret, "BTC Log-Returns (%)")
```

**Interpretation:** BTC log-returns have a small positive mean return (0.146%) 
but substantial volatility (SD = 3.53%), with the distribution showing negative 
skewness (−0.67) and very high kurtosis (12.25), indicating that large negative 
shocks and extreme movements occur more often than under normality. This is also
evident in the QQ plot, where both tails deviate notably from the reference
line, confirming a heavy-tailed, non-normal distribution. The time-series plot
suggests volatility clustering, with calm periods followed by turbulent periods,
which is consistent with conditional heteroskedasticity and supports the use of 
GARCH-type models. At the same time, the ACF and PACF of returns show only weak 
serial correlation at most lags, implying limited linear dependence in the mean 
process. Furthermore, the stationarity tests agree that BTC log-returns are 
stationary: the ADF test rejects the unit root null, while the KPSS test fails 
to reject stationarity, making the series suitable for return-based models such 
as ARMA, VAR, and GARCH. Finally, the ARCH LM test strongly rejects the null 
hypothesis of no ARCH effects indicating significant volatility clustering and 
supporting the use of GARCH-type models to capture time-varying variance in BTC 
returns.

```{r eda-btc-log-price}
eda(btc_log, "BTC Log-Price")
```

**Interpretation:**
BTC log-price shows a strong upward trend over time, reflecting the long-term 
growth of the asset, with values ranging from about 5.90 to 11.73 and an average
of 9.50. The distribution is slightly negatively skewed (−0.68), indicating 
somewhat more mass in the upper range of prices relative to the lower range, 
while the kurtosis (−0.39) suggests lighter tails compared with a normal 
distribution, which is typical for trending level series rather than returns. 
The QQ plot confirms deviations from normality, particularly due to the trend 
structure in the data. The ACF displays very high persistence across many lags,
while the PACF shows a large spike at the first lag followed by near-zero 
values, both of which are characteristic of a nonstationary time series with a 
stochastic trend. The formal stationarity tests support this: the ADF test fails
to reject the unit root null (p = 0.485), and the KPSS test rejects the null of
stationarity (p = 0.01). Together, these results indicate that BTC log-price is 
nonstationary and likely contains a unit root, meaning differencing or 
state-space trend modeling (e.g., SARIMA differencing or BSTS trend components) 
is required before forecasting.Finally, the ARCH LM test strongly rejects the 
null hypothesis of no ARCH effects indicating significant volatility clustering 
and supporting the use of GARCH-type models to capture time-varying variance in 
BTC price.

## Correlation Structure

```{r correlation-heatmap}
# Build dataframe for correlation analysis
pBTC_c <- log(daily_data$BTC)
rBTC_c <- diff(log(daily_data$BTC)) * 100
rSPY_c <- diff(log(daily_data$SPY)) * 100
rQQQ_c <- diff(log(daily_data$QQQ)) * 100
dFF_c  <- diff(daily_data$FedFunds)
dGS10_c <- diff(daily_data$DGS10)
dCPI_c  <- diff(log(daily_data$CPI)) * 100
dUNR_c  <- diff(daily_data$Unemployment)

corr_df <- na.omit(data.frame(
  pBTC = pBTC_c,
  rBTC = rBTC_c,
  rSPY = rSPY_c,
  rQQQ = rQQQ_c,
  dFF = dFF_c,
  dGS10 = dGS10_c,
  dCPI = dCPI_c,
  dUNR = dUNR_c
))

cor_mat <- cor(corr_df)

# Melt for ggplot heatmap
cor_long <- reshape2::melt(cor_mat)

ggplot(cor_long, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 3) +
  scale_fill_gradient2(low = "steelblue", mid = "white", high = "tomato",
                       midpoint = 0, limits = c(-1, 1)) +
  theme_minimal() +
  labs(title = "Pearson Correlation — Log BTC + Return/Change Series",
       x = "", y = "", fill = "Corr")
```

**Key observations:** The correlation matrix shows that BTC returns have 
moderate positive correlations with equity market returns, particularly SPY and 
QQQ (≈0.24), suggesting that BTC sometimes moves with broader risk assets. As 
expected, SPY and QQQ are extremely highly correlated (≈0.93) because both 
represent large-cap U.S. equities with heavy overlap in holdings. In contrast, 
correlations between BTC returns and macroeconomic variables such as Fed Funds, 
the 10-year Treasury yield (DGS10), CPI changes, and unemployment changes are 
very close to zero, indicating little contemporaneous linear relationship. The 
BTC log-price itself also shows near-zero correlations with these differenced 
macro variables, which is consistent with the fact that the macro series 
represent changes while BTC log-price is a level series. Overall, the heatmap
suggests that BTC is more connected to equity market movements than to short-run 
macroeconomic fluctuations, though the relationships remain relatively weak 
outside of the equity indices.


# Data Processing

## Anomaly Detection

```{r anomaly-detection-log-btc-returns}
# Flag extreme BTC return observations (beyond 3 SD)
mu  <- mean(btc_ret, na.rm = TRUE)
sig <- sd(btc_ret, na.rm = TRUE)

outliers <- btc_ret[abs(btc_ret - mu) > 3 * sig]

cat("Observations beyond ±3σ:", length(outliers), "\n")
cat("Dates of extreme moves:\n")
print(head(outliers, 20))

# Plot with outlier threshold
plot(as.numeric(btc_ret),
     type = "l",
     main = "BTC Log-Returns — Outlier Detection (±3σ)",
     col = "steelblue",
     lwd = 0.8)

abline(h = c(mu + 3*sig, mu - 3*sig),
       col = "red",
       lty = 2,
       lwd = 2)

legend("topright",
       legend = "±3σ threshold",
       col = "red",
       lty = 2,
       lwd = 2)
```

**Decision:** The analysis identifies 65 observations beyond the ±3σ threshold, 
indicating the presence of extreme outliers in BTC returns. However, these 
observations are retained because, in financial time series, large price 
movements typically reflect genuine market events—such as regulatory 
announcements, liquidity shocks, exchange failures, or sudden shifts in investor
sentiment—rather than data errors. Removing them would distort the true 
distribution of returns and understate the volatility and risk inherent in the 
asset. Their presence is consistent with the heavy-tailed behavior observed in 
the EDA, and the Student-t distribution used in the GARCH model is specifically
designed to accommodate such extreme observations.

## Missing Value Summary

```{r missing-summary}
na_counts <- colSums(is.na(daily_data))
cat("Missing values per series AFTER imputation:\n")
print(na_counts)
```


# Feature Engineering

## Stationary Transformations

```{r feature-engineering}
# All series transformed to stationarity before modeling
pBTC  <- log(daily_data$BTC)
rBTC  <- diff(log(daily_data$BTC))   * 100   # log-return (%)
rSPY  <- diff(log(daily_data$SPY))   * 100   # log-return (%)
rQQQ  <- diff(log(daily_data$QQQ))   * 100   # log-return (%)
dFF   <- diff(daily_data$FedFunds)           # 1st-diff of rate level
dGS10 <- diff(daily_data$DGS10)              # 1st-diff of rate level
dCPI  <- diff(log(daily_data$CPI))   * 100   # log-return (%)
dUNR  <- diff(daily_data$Unemployment)       # 1st-diff of rate level

var_df <- na.omit(merge(pBTC, rBTC, rSPY, rQQQ, dFF, dGS10, dCPI, dUNR))
colnames(var_df) <- c("pBTC", "rBTC","rSPY","rQQQ","dFF","dGS10","dCPI","dUNR")

cat("Feature matrix dimensions:", dim(var_df), "\n")
head(var_df)
```

**Feature engineering rationale:**

| Raw Variable | Transformation | Reason |
|-------------|---------------|--------|
| BTC | Log | Preserves the level-series interpretation for price modeling while helping stabilize variance |
| BTC, SPY, QQQ prices; CPIAUCSL | Log-difference × 100 | Converts prices and the CPI index into percentage changes, helping remove trend/nonstationarity and making the transformed series easier to compare across variables |
| DFF, DGS10, UNRATE | First difference | Converts levels into changes, which are more likely to be stationary than the original series |

## EDA of Regressors

```{r eda_spy}
# EDA for SPY
eda(rSPY,  "SPY Log-Returns (%)")
```

```{r eda_qqq}
# EDA for QQQ
eda(rQQQ,  "QQQ Log-Returns (%)")
```

```{r eda_ff}
# EDA for FF
eda(dFF,   "Fed Funds Change")
```

```{r eda_gs10}
# EDA for GS10
eda(dGS10, "10-Year Treasury Change")
```

```{r eda_cpi}
# EDA for CPI
eda(dCPI,  "Inflation (Log-Diff %)")
```

```{r eda_unr}
# EDA for UNR
eda(dUNR,  "Unemployment Change")
```

**Interpretation:** The equity return series (rSPY and rQQQ) strongly reject the
unit root null in the ADF test and fail to reject stationarity in the KPSS test,
confirming that the log-return transformation successfully removes 
nonstationarity. Similar results hold for the macroeconomic variables after 
differencing or log-differencing: ADF tests reject the presence of a unit root 
while KPSS tests generally support stationarity, suggesting the transformations
effectively remove persistent trends.

## Train / Test Split

```{r train-test-split}
train_xts <- var_df[index(var_df) < test_start]
train_xts <- train_xts[, colnames(train_xts) != "pBTC"]

test_xts  <- var_df[index(var_df) >= test_start]
test_xts  <- test_xts[, colnames(test_xts) != "pBTC"]

cat("Train range:", as.character(start(index(train_xts))),
    "to", as.character(end(index(train_xts))), "\n")
cat("Train size:", nrow(train_xts), "observations\n")
cat("Test  range:", as.character(start(index(test_xts))),
    "to", as.character(end(index(test_xts))), "\n")
cat("Test  size:", nrow(test_xts), "observations\n")

train_Y <- as.matrix(train_xts)
test_Y  <- as.matrix(test_xts)
```

**Note:** The test set (June 2025 onward) is a true hold-out; no model is tuned 
on test data. The rolling 1-step-ahead evaluation in the VAR+GARCH section 
re-estimates the model at each step using only data available up to that point 
— eliminating look-ahead bias.


# Proposed Models

This section presents four models in order of complexity. Each is justified by 
both theoretical motivation and empirical diagnostics.

## Model 1: ARMA — Baseline Mean Model

The ARMA baseline tests whether BTC returns contain any linear autocorrelation 
structure in their **mean**. It serves as the mean equation for the GARCH model.

```{r arma-fit}
arma_fit <- auto.arima(as.numeric(btc_ret),
                       d = 0,
                       stationary = TRUE,
                       seasonal   = FALSE,
                       max.p = 3, max.q = 3)

cat("Selected ARMA order:", arma_fit$arma[1], arma_fit$arma[2], "\n")
summary(arma_fit)
```
**Key observations:**
An ARMA(0,2) model (i.e., ARIMA(0,0,2) with a non-zero mean) was selected for 
the BTC log-returns, meaning returns are modeled as a constant mean plus a 
moving-average component driven by shocks from the previous two days. The 
estimated mean return is 0.1456% (s.e. 0.0585), while the MA coefficients are
small (ma1 = −0.0282, s.e. 0.0165; ma2 = 0.0302, s.e. 0.0163), suggesting only 
weak short-run dependence in the return series. Model fit statistics are 
AIC = 19589.5 (AICc = 19589.51) and BIC = 19614.31, with estimated innovation 
variance σ² = 12.47 (log-likelihood −9790.75). On the training set, forecast 
errors are RMSE = 3.53 and MAE = 2.30, and the residual autocorrelation at lag 1
is essentially zero (ACF1 ≈ 0.00008), indicating the model leaves little linear
autocorrelation in the mean process.


## Model 2: GARCH — Volatility Modeling

### ARCH Test (Pre-condition)

```{r arch-test}
# ARCH test: H0 = no ARCH effects
cat("ARCH Test on raw BTC returns:\n")
ArchTest(as.numeric(btc_ret), lags = 12)
```

**Interpretation:** A significant ARCH test (p < 0.05) confirms volatility 
clustering — large shocks tend to be followed by large shocks. This validates
fitting a GARCH model.

### ARCH(1) — Benchmark Variance Model

```{r arch1-fit}
spec_arch <- ugarchspec(
  variance.model     = list(model = "sGARCH", garchOrder = c(1, 0)),
  mean.model         = list(armaOrder = c(0, 2), include.mean = TRUE),
  distribution.model = "norm"
)

fit_arch <- ugarchfit(spec = spec_arch, data = btc_ret)
show(fit_arch)
```

```{r arch1-diagnostics}
z_arch <- residuals(fit_arch, standardize = TRUE)

cat("Ljung-Box on standardized residuals (serial correlation):\n")
print(Box.test(as.numeric(z_arch), lag = 20, type = "Ljung-Box"))

cat("\nARCH test on standardized residuals (remaining ARCH effects):\n")
print(ArchTest(as.numeric(z_arch), lags = 12))

Acf(as.numeric(z_arch)^2,
    main = "ACF of Squared Std Residuals — ARCH(1)\n(should be insignificant if adequate)")
```
**Interpretation:**
The Ljung–Box test on standardized residuals strongly rejects the null 
hypothesis of no serial correlation (p < 2.2e-16), indicating that some linear
dependence remains in the residuals and the mean specification may not fully 
capture all temporal dynamics. However, the ARCH LM test on standardized
residuals yields a p-value of 1, suggesting that the GARCH component 
successfully removes remaining ARCH effects and adequately models conditional 
heteroskedasticity.

### GARCH(1,1) — Full Volatility Model

```{r garch11-fit}
spec_garch <- ugarchspec(
  variance.model     = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model         = list(armaOrder = c(0, 2), include.mean = TRUE),
  distribution.model = "std"   # Student-t to handle heavy tails
)

fit_garch <- ugarchfit(spec_garch, data = btc_ret)
show(fit_garch)
```

```{r garch11-diagnostics}
z_garch <- residuals(fit_garch, standardize = TRUE)

cat("Ljung-Box on standardized residuals:\n")
print(Box.test(as.numeric(z_garch), lag = 20, type = "Ljung-Box"))

cat("\nARCH test on standardized residuals:\n")
print(ArchTest(as.numeric(z_garch), lags = 12))

Acf(as.numeric(z_garch)^2,
    main = "ACF of Squared Std Residuals — GARCH(1,1)\n(should be insignificant if adequate)")
```

**Interpretation:**
Compared with the earlier ARCH specification, the GARCH model provides a better 
overall fit. In the ARCH model, the Ljung–Box test on standardized residuals 
produced an extremely small p-value (p < 2.2e-16), indicating strong remaining
serial correlation and suggesting that the model did not adequately capture the 
underlying dynamics of the series. In the GARCH model, the Ljung–Box statistic 
is less small (p = 0.00136), showing that the remaining autocorrelation is
substantially reduced. Additionally, the ARCH LM test yields a high p-value 
(p = 0.8575), meaning we fail to reject the null hypothesis of no remaining ARCH
effects. This indicates that the GARCH specification successfully captures 
volatility clustering and improves the overall modeling of the conditional 
variance relative to the simpler ARCH model.

**Justification for GARCH(1,1) over ARCH(1):**
- The ARCH(1) model only allows one lag of squared residuals to drive current 
variance
- The GARCH(1,1) adds a lagged variance term, providing a longer "memory" of 
past volatility — well-suited for the persistent volatility clusters observed in
BTC
- Student-t distribution captures heavy tails better than Gaussian



## Model 3: VAR — Multivariate Mean Forecasting

VAR models the joint dynamics of all 7 stationary series, allowing each 
variable's past values to help predict BTC returns. This tests H2 (macro Granger
causality).

### Visualize Training Series

```{r var-eda-plot}
tmp_df   <- data.frame(Date = index(train_xts), coredata(train_xts))
tmp_long <- reshape2::melt(tmp_df, id.vars = "Date",
                           variable.name = "Series", value.name = "Value")

ggplot(tmp_long, aes(x = Date, y = Value)) +
  geom_line(alpha = 0.6, color = "steelblue") +
  facet_wrap(~Series, scales = "free_y", ncol = 1) +
  theme_minimal() +
  labs(title = "All VAR Input Series — Training Set",
       subtitle = "All series are stationary (differenced or log-differenced)",
       x = "Date", y = "")
```

### ACF / PACF

```{r var-acf-pacf}
par(mfrow = c(2, 1))
Acf(var_df$rBTC,  lag.max = 60, main = "ACF: rBTC")
Pacf(var_df$rBTC, lag.max = 60, main = "PACF: rBTC")
par(mfrow = c(1, 1))
```

### Cross-Correlation (Predictors → BTC)

```{r var-ccf}
par(mfrow = c(3, 2))
for (xnm in setdiff(colnames(var_df), "rBTC")) {
  ccf(as.numeric(var_df[, xnm]), as.numeric(var_df$rBTC),
      lag.max = 60,
      main    = paste("CCF:", xnm, "→ rBTC"),
      ylab    = "Cross-correlation")
}
par(mfrow = c(1, 1))
```

**Interpretation:** 
The cross-correlation function (CCF) plots examine whether lagged values of 
financial and macroeconomic variables contain predictive information for Bitcoin
returns. Most correlations remain within the 95% confidence bounds, indicating 
that the lead-lag relationships between BTC returns and variables such as SPY,
QQQ, interest rates, inflation, and unemployment are generally weak. A few 
isolated spikes appear at certain lags, particularly for unemployment and 
inflation, suggesting occasional delayed associations, but no consistent pattern
emerges across the predictors. Overall, the CCF analysis suggests that Bitcoin
returns exhibit limited short-term dependence on these external variables, 
although they may still contribute to the multivariate dynamics captured in the
subsequent VAR modeling framework.

### Lag Order Selection

```{r var-lag-order}
ord <- VARorder(train_Y, maxp = 10, output = TRUE)

p <- ord$aicor[1]  # AIC-selected lag order
cat("AIC-selected lag order p =", p, "\n")
```

### Granger Causality Tests

```{r granger-tests}
cat("Pairwise Granger Causality Tests (alpha = 0.05)\n")
cat("H0: predictor does NOT Granger-cause rBTC\n\n")

alpha <- 0.05
for (cause_var in setdiff(colnames(train_xts), "rBTC")) {
  formula_obj <- as.formula(paste("rBTC ~", cause_var))
  test_result <- grangertest(formula_obj, order = p, data = train_xts)
  pval        <- test_result$`Pr(>F)`[2]
  verdict     <- if (pval < alpha) "GRANGER-CAUSES" else "does NOT cause"

  cat(sprintf("%-6s -> rBTC: F = %6.3f, p = %7.4f => %s\n",
              cause_var, test_result$F[2], pval, verdict))
}
```
**Interpretation:**
These results report pairwise Granger causality tests evaluating whether other 
financial and macroeconomic variables help predict Bitcoin returns (rBTC). The 
null hypothesis for each test is that the predictor does not Granger-cause rBTC,
meaning that its past values do not provide additional predictive information 
beyond the past values of Bitcoin returns themselves. Using a significance level 
of α = 0.05, all p-values are well above 0.05. This means we fail to reject the 
null hypothesis for every predictor, including equity market returns and 
macroeconomic variables. In other words, based on this test and the selected lag
structure, none of these variables provide statistically significant predictive
power for Bitcoin returns. This result suggests that BTC returns are largely 
driven by their own past dynamics rather than short-term spillovers from 
traditional financial markets or macroeconomic indicators, which is consistent 
with the weak lead–lag relationships observed in the earlier cross-correlation
analysis. Nonetheless, we will fit VAR to confirm this.

### VAR Model Fit

```{r var-fit, echo=TRUE, results='hide'}
var_fit <- MTS::VAR(train_Y, p = p, include.mean = TRUE)

cat("Residual Covariance Matrix (Sigma):\n")
print(round(var_fit$Sigma, 6))
```

### VAR Residual Diagnostics

```{r var-mq-test}
# Multivariate Ljung-Box (Portmanteau) test on VAR residuals
mq(var_fit$residuals, lag = 10)
```
**Interpretation:**
The Ljung–Box test results evaluate whether there is remaining serial 
correlation in the residuals of the fitted VAR model. The null hypothesis of the
test is that the residuals are independently distributed (no autocorrelation) up
to lag m. Across all tested lags from m=1 to m=10, the reported p-values are 
equal to 1, which are far above the 0.05 significance threshold. This means we 
fail to reject the null hypothesis at every lag, indicating that there is no 
evidence of residual autocorrelation remaining in the model errors. The 
accompanying plot confirms this result, as all p-values lie well above the 
significance boundary (blue dashed line). Overall, these diagnostics suggest 
that the VAR model with the selected lag structure adequately captures the 
temporal dependencies in the data, leaving residuals that behave approximately
like white noise, which is a desirable property for a well-specified time series
model. This is reflected below in the residuals plot below as well. 

```{r var-residual-plots}
residuals_var <- var_fit$residuals
colnames(residuals_var) <- colnames(train_Y)

par(mfrow = c(4, 2), mar = c(2, 3, 2, 1))
for (i in 1:7) {
  plot(residuals_var[, i], type = "l",
       main = paste("VAR Residuals:", colnames(residuals_var)[i]),
       ylab = "", xlab = "")
  abline(h = 0, lty = 2, col = "red")
}
par(mfrow = c(1, 1))
```

```{r var-residual-qq}
par(mfrow = c(4, 2), mar = c(2, 3, 2, 1))
for (i in 1:7) {
  qqnorm(residuals_var[, i],
         main = paste("QQ Plot:", colnames(residuals_var)[i]))
  qqline(residuals_var[, i], col = "red")
}
par(mfrow = c(1, 1))
```
**Interpretation:**
The QQ plots show the distribution of the VAR model residuals for each variable 
compared with a theoretical normal distribution. If the residuals were perfectly
normal, the points would lie along the red reference line. While the residuals 
for most variables align closely with the line in the center of the distribution, 
several plots exhibit noticeable deviations in the tails, particularly for BTC 
and equity return residuals, where extreme observations fall away from the line. 
This indicates the presence of heavy tails and occasional outliers, meaning the 
residuals are not perfectly normally distributed. However, the relatively good 
alignment in the middle suggests the model captures the main dynamics of the 
series reasonably well, while the tail deviations reflect the well-known
tendency of financial returns to produce extreme shocks more frequently than a 
normal distribution would predict.

### VAR Multi-Step Forecast (Test Set)

```{r var-multistep-forecast, echo=TRUE, results='hide'}
h <- nrow(test_Y)

# h-step ahead forecast from end of training
fc       <- VARpred(var_fit, h = h)
pred_mat <- fc$pred
colnames(pred_mat) <- colnames(train_Y)

actual_rBTC <- test_Y[, "rBTC"]
pred_rBTC   <- pred_mat[, "rBTC"]

rmse_var <- sqrt(mean((actual_rBTC - pred_rBTC)^2))
mae_var  <- mean(abs(actual_rBTC - pred_rBTC))
```

```{r var-accuracy}
cat("VAR Multi-Step Forecast Accuracy (rBTC):\n")
cat(sprintf("  RMSE: %.4f\n  MAE:  %.4f\n", rmse_var, mae_var))
```


## Model 4: VAR + GARCH — Rolling 1-Step Forecast

Multi-step VAR forecasts degrade quickly as errors compound. The rolling 
1-step-ahead approach re-estimates both the VAR (mean) and GARCH (variance) at 
each time step, using all data available up to that point.

### Static VAR + GARCH Diagnostics

```{r var-garch-static}
# Re-fit VAR on full training set
var_fit <- MTS::VAR(train_Y, p = p, include.mean = TRUE)
colnames(var_fit$residuals) <- colnames(train_Y)

resid_BTC <- na.omit(as.numeric(var_fit$residuals[, "rBTC"]))

cat("ARCH Test on VAR residuals (BTC equation):\n")
ArchTest(resid_BTC, lags = 12)

spec_var_garch <- ugarchspec(
  variance.model     = list(model = "sGARCH", garchOrder = c(1, 1)),
  mean.model         = list(armaOrder = c(0, 0), include.mean = FALSE),
  distribution.model = "std"
)

fit_var_garch <- ugarchfit(spec = spec_var_garch, data = resid_BTC)
show(fit_var_garch)

z_vg <- residuals(fit_var_garch, standardize = TRUE)
cat("\nLjung-Box on VAR+GARCH standardized residuals:\n")
print(Box.test(as.numeric(z_vg), lag = 20, type = "Ljung-Box"))
cat("\nARCH test on VAR+GARCH standardized residuals:\n")
print(ArchTest(as.numeric(z_vg), lags = 12))

Acf(as.numeric(z_vg)^2,
    main = "ACF: Squared Std Residuals (VAR+GARCH)\n(should be insignificant if adequate)")
```
**Interpretation:**
The Ljung–Box test on the standardized residuals has a p-value of 0.141, so we
fail to reject the null hypothesis of no autocorrelation in the residuals. This 
suggests that the VAR model adequately captures the mean dynamics of the series. 
In addition, the ARCH LM test yields a p-value of 0.85, meaning we also fail to 
reject the null hypothesis of no remaining ARCH effects. This indicates that the 
GARCH(1,1) specification has successfully captured the volatility clustering in
the residuals. The squared residual ACF plot supports this conclusion, as it 
shows no significant spikes. Overall, these diagnostics suggest that the static 
VAR + GARCH specification is reasonably well specified, with residuals that
behave approximately like white noise.

### Rolling 1-Step-Ahead Evaluation

```{r var-garch-rolling, message=FALSE, warning=FALSE, echo=TRUE, results='hide'}
h <- nrow(test_Y)

pred_mean_rBTC  <- rep(NA, h)
pred_sigma_rBTC <- rep(NA, h)

rolling_Y <- train_Y

for (i in 1:h) {
  # 1) Fit VAR on expanding window
  var_i <- MTS::VAR(rolling_Y, p = p, include.mean = TRUE)

  # 2) 1-step mean forecast
  fc_i     <- VARpred(var_i, h = 1)
  pred_vec <- fc_i$pred
  pred_mat1 <- if (is.null(dim(pred_vec))) matrix(pred_vec, nrow = 1) else as.matrix(pred_vec)
  colnames(pred_mat1) <- colnames(rolling_Y)

  pred_mean_rBTC[i] <- as.numeric(pred_mat1[1, "rBTC"])

  # 3) GARCH(1,1) on BTC residuals from VAR
  res_mat <- var_i$residuals
  if (is.null(colnames(res_mat))) colnames(res_mat) <- colnames(rolling_Y)

  resid_BTC_i <- na.omit(as.numeric(res_mat[, "rBTC"]))
  garch_i     <- ugarchfit(spec = spec_var_garch, data = resid_BTC_i, solver = "hybrid")
  gfc_i       <- ugarchforecast(garch_i, n.ahead = 1)

  pred_sigma_rBTC[i] <- as.numeric(sigma(gfc_i)[1])

  # 4) Expand window with actual observation
  rolling_Y <- rbind(rolling_Y, test_Y[i, ])
}

actual_rBTC <- as.numeric(test_Y[, "rBTC"])

rmse_rolling <- sqrt(mean((actual_rBTC - pred_mean_rBTC)^2))
mae_rolling  <- mean(abs(actual_rBTC - pred_mean_rBTC))

cat("Rolling 1-Step VAR+GARCH Forecast Accuracy (rBTC):\n")
cat(sprintf("  RMSE: %.4f\n  MAE:  %.4f\n", rmse_rolling, mae_rolling))

vol_rmse <- sqrt(mean((actual_rBTC^2 - pred_sigma_rBTC^2)^2))
cat(sprintf("  Volatility proxy RMSE (r² vs σ²): %.4f\n", vol_rmse))
```

**Interpretation:**
In the rolling 1-step-ahead VAR+GARCH evaluation, the VAR model fit on the 
expanding training window shows strong overall in-sample fit with information 
criteria values of AIC = −21.03, BIC = −20.28, and HQ = −20.77, which are 
primarily useful for comparing alternative lag specifications. The forecast 
output reports the one-step-ahead predicted means for each variable at the 
forecast origin, with associated prediction uncertainty: BTC has much larger 
forecast standard error and RMSE (SE ≈ 3.51, RMSE ≈ 3.54) than SPY/QQQ and the 
macro series, reflecting BTC’s substantially higher intrinsic volatility. 
Out-of-sample performance for the key target, Bitcoin returns, is summarized by 
the rolling forecast accuracy metrics: the VAR+GARCH framework achieves an RMSE 
of 1.93 and MAE of 1.45 for rBTC, indicating that the average one-day-ahead 
forecast error is about 1.45 percentage points. Finally, the volatility proxy 
RMSE comparing realized squared returns r^2 to the GARCH-implied variance σ^2 is
6.62, highlighting that while the model captures time-varying volatility, 
realized volatility remains noisy and difficult to predict precisely at the 
daily horizon.


## Model 5: SARIMA — Univariate Log-Price Baseline

SARIMA serves as a principled univariate baseline for log-price (level)
forecasting.Weekly seasonality (frequency = 7) is specified to capture 
day-of-week patterns in BTC trading volume and price.

```{r sarima-data}
btc_log   <- log(daily_data$BTC)
btc_log   <- btc_log[!is.na(btc_log)]

btc_train <- btc_log[index(btc_log) < test_start]
btc_test  <- btc_log[index(btc_log) >= test_start]

cat("SARIMA Train: n =", length(btc_train), "\n")
cat("SARIMA Test:  n =", length(btc_test),  "\n")
```

```{r sarima-fit}
y_train <- ts(as.numeric(btc_train), frequency = 7)
y_test  <- ts(as.numeric(btc_test),  frequency = 7)

fit_sarima <- auto.arima(y_train, seasonal = TRUE)
summary(fit_sarima)

h           <- length(y_test)
fc_sarima   <- forecast(fit_sarima, h = h)
```

```{r sarima-accuracy}
rmse_sarima <- sqrt(mean((as.numeric(y_test) - as.numeric(fc_sarima$mean))^2))
mae_sarima  <- mean(abs(as.numeric(y_test)   - as.numeric(fc_sarima$mean)))

cat("SARIMA Forecast Accuracy (log-price scale):\n")
cat(sprintf("  RMSE: %.4f\n  MAE:  %.4f\n", rmse_sarima, mae_sarima))
```

**Note:** SARIMA RMSE/MAE are on the **log-price scale**, not the returns scale.
They are not directly comparable to VAR+GARCH metrics, which are on the returns
scale. We use log prices for the SARIMA model because ARIMA-type models 
typically require the series to be non-stationary in levels but stationary after
differencing, and taking the log stabilizes variance and allows the model to 
capture price trends and proportional changes more effectively. In contrast, the
VAR+GARCH framework is designed to model stationary return series and 
time-varying volatility. 


**Interpretation:**
The SARIMA model fitted to the log Bitcoin price series is an SARIMA(0,1,2) with 
drift, meaning the series is first-differenced to remove nonstationarity while 
two moving-average terms capture short-term shocks and the drift term represents 
a small underlying upward trend in the log price. The estimated MA coefficients 
(−0.0293 and 0.0316) are relatively small, suggesting that only modest short-run 
error dependence exists after differencing. The model achieves strong in-sample 
fit, with a low residual variance (σ² ≈ 0.0013) and information criteria values 
of AIC = −13077.23 and BIC = −13052.66. Training error metrics are also small 
(RMSE ≈ 0.036, MAE ≈ 0.0236), and the near-zero ACF1 of the residuals indicates 
little remaining autocorrelation, suggesting the model captures most of the 
serial structure in the differenced log-price series. When evaluated on the test 
set, the model produces an out-of-sample forecast RMSE of 0.2416 and MAE of 
0.1665 on the log-price scale, indicating reasonable predictive accuracy for 
short-term Bitcoin price movements.


## Model 6: BSTS — Bayesian Structural Time Series

BSTS decomposes log-price into trend + seasonal + regression components. Unlike 
SARIMA, BSTS explicitly models the structural components and incorporates 
macroeconomic regressors with spike-and-slab priors that perform automatic
variable selection.

```{r bsts-data}
X <- var_df[, c("rSPY","rQQQ","dFF","dGS10","dCPI","dUNR")]

# Align log-BTC with predictor dates
y <- btc_log[index(btc_log) %in% index(X)]
X <- X[index(X) %in% index(y)]
stopifnot(all(index(y) == index(X)))

y_train_b <- as.numeric(y[index(y) < test_start])
y_test_b  <- as.numeric(y[index(y) >= test_start])

X_train <- as.matrix(X[index(X) < test_start])
X_test  <- as.matrix(X[index(X) >= test_start])

cat("BSTS Train n =", length(y_train_b), " | Test n =", length(y_test_b), "\n")
```

```{r bsts-fit, message=FALSE}
ss <- list()
ss <- AddLocalLevel(ss,   y_train_b)
ss <- AddSeasonal(ss,     y_train_b, nseasons = 7)

set.seed(123)
fit_bsts <- bsts(
  formula             = y_train_b ~ X_train,
  state.specification = ss,
  niter               = 2000,
  ping                = 0
)

summary(fit_bsts)
plot(fit_bsts, main = "BSTS: State Components + Coefficient Posteriors")
```

```{r bsts-forecast}
pred_bsts     <- predict(fit_bsts, newdata = X_test)
pred_log_bsts <- as.numeric(pred_bsts$mean)

rmse_bsts <- sqrt(mean((y_test_b - pred_log_bsts)^2))
mae_bsts  <- mean(abs(y_test_b  - pred_log_bsts))

cat("BSTS Forecast Accuracy (log-price scale):\n")
cat(sprintf("  RMSE: %.4f\n  MAE:  %.4f\n", rmse_bsts, mae_bsts))
```
**Interpretation:**
The BSTS model produces out-of-sample forecast errors of RMSE = 0.1056 and 
MAE = 0.0901 on the log-price scale, indicating relatively strong predictive 
performance for Bitcoin prices. Compared to the SARIMA model (RMSE = 0.2416, 
MAE = 0.1665), the BSTS approach yields substantially lower forecast errors, 
suggesting that incorporating a Bayesian structural framework with latent trend 
components better captures the underlying dynamics of the log-price series. 
Because BSTS models explicitly decompose the series into components such as 
trend, noise, and potentially seasonal or regression effects, they can adapt 
more flexibly to structural changes in the data, which is particularly useful 
for highly volatile assets like Bitcoin. As with the SARIMA model, these 
accuracy metrics are measured on the log-price scale, meaning they are not 
directly comparable to the VAR+GARCH results, which are evaluated on the 
returns scale. We don't use log returns because BSTS deomposes a series into 
latent components including trend. 



# Results

## Plot Helper Functions

```{r plot-helpers}
# ── Unified forecast plot (any scale) ─────────────────────────────────────────
plot_forecast <- function(dates, actual, predicted, title,
                          y_label = "Value", rmse = NULL, mae = NULL) {
  subtitle_txt <- if (!is.null(rmse) && !is.null(mae))
    paste0("Test period  |  RMSE = ", round(rmse, 2), "  |  MAE = ", round(mae, 2))
  else ""

  df <- data.frame(Date = dates, Actual = actual, Predicted = predicted)
  df_long <- reshape2::melt(df, id.vars = "Date",
                            variable.name = "Series", value.name = "Value")

  ggplot(df_long, aes(x = Date, y = Value, colour = Series)) +
    geom_line(linewidth = 0.85, alpha = 0.85) +
    scale_colour_manual(values = c("Actual" = "#2166ac", "Predicted" = "#d6604d")) +
    scale_y_continuous(labels = scales::comma) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "top",
          plot.subtitle    = element_text(size = 10, colour = "grey40"),
          panel.grid.minor = element_blank()) +
    labs(title = title, subtitle = subtitle_txt,
         x = "Date", y = y_label, colour = "")
}

# ── Convert VAR/GARCH return forecasts → USD price ────────────────────────────
# Starting from last_price, chain: P_t = P_{t-1} * exp(r_t / 100)
returns_to_price <- function(returns_pct, last_price) {
  prices <- numeric(length(returns_pct))
  prices[1] <- last_price * exp(returns_pct[1] / 100)
  for (i in seq_along(returns_pct)[-1])
    prices[i] <- prices[i - 1] * exp(returns_pct[i] / 100)
  prices
}
```

## Actual BTC Price Plots — All Models (USD Scale)

All four models are converted to **USD price** for direct visual comparison.

- **VAR / VAR+GARCH:** return forecasts are compounded forward from the last training
  price using `P_t = P_{t-1} × exp(r_t / 100)`.
- **SARIMA / BSTS:** log-price forecasts are back-transformed with `exp(·)`.

RMSE and MAE in the subtitles are also on the **USD scale**, making them directly
comparable across all models for the first time.

```{r usd-prep}
# Last known BTC price at end of training window
last_train_price <- as.numeric(tail(daily_data$BTC[index(daily_data$BTC) < test_start], 1))
cat("Last training price (USD):", scales::dollar(last_train_price), "\n")

# Actual BTC price during the test window
test_dates_btc <- index(daily_data$BTC[index(daily_data$BTC) >= test_start])
actual_usd     <- as.numeric(daily_data$BTC[index(daily_data$BTC) >= test_start])

# ── VAR multi-step → USD ──────────────────────────────────────────────────────
var_dates      <- index(test_xts)
pred_usd_var   <- returns_to_price(as.numeric(pred_mat[, "rBTC"]), last_train_price)
actual_usd_var <- as.numeric(daily_data$BTC[as.character(var_dates)])

rmse_usd_var <- sqrt(mean((actual_usd_var - pred_usd_var)^2, na.rm = TRUE))
mae_usd_var  <- mean(abs(actual_usd_var  - pred_usd_var),    na.rm = TRUE)

# ── VAR+GARCH rolling → USD ───────────────────────────────────────────────────
pred_usd_vg   <- returns_to_price(pred_mean_rBTC, last_train_price)
actual_usd_vg <- actual_usd_var   # same test dates

rmse_usd_vg <- sqrt(mean((actual_usd_vg - pred_usd_vg)^2, na.rm = TRUE))
mae_usd_vg  <- mean(abs(actual_usd_vg  - pred_usd_vg),    na.rm = TRUE)

# ── SARIMA → USD (exp of log-price forecast) ──────────────────────────────────
sarima_dates   <- as.Date(index(btc_test))
pred_usd_sar   <- exp(as.numeric(fc_sarima$mean))
actual_usd_sar <- exp(as.numeric(btc_test))

rmse_usd_sar <- sqrt(mean((actual_usd_sar - pred_usd_sar)^2, na.rm = TRUE))
mae_usd_sar  <- mean(abs(actual_usd_sar  - pred_usd_sar),    na.rm = TRUE)

# ── BSTS → USD (exp of log-price forecast) ───────────────────────────────────
bsts_dates    <- index(X[index(X) >= test_start])
pred_usd_bsts <- exp(pred_log_bsts)
actual_usd_bsts <- exp(y_test_b)

rmse_usd_bsts <- sqrt(mean((actual_usd_bsts - pred_usd_bsts)^2, na.rm = TRUE))
mae_usd_bsts  <- mean(abs(actual_usd_bsts  - pred_usd_bsts),    na.rm = TRUE)
```

### Model 3: VAR — Multi-Step Ahead (USD)

```{r plot-var-usd}
plot_forecast(
  dates     = var_dates,
  actual    = actual_usd_var,
  predicted = pred_usd_var,
  title     = "Model 3: VAR — Multi-Step Ahead",
  y_label   = "BTC Price (USD)",
  rmse      = rmse_usd_var,
  mae       = mae_usd_var
)
```

### Model 4: VAR+GARCH — Rolling 1-Step Ahead (USD)

```{r plot-vargarch-usd}
plot_forecast(
  dates     = var_dates,
  actual    = actual_usd_vg,
  predicted = pred_usd_vg,
  title     = "Model 4: VAR+GARCH — Rolling 1-Step Ahead",
  y_label   = "BTC Price (USD)",
  rmse      = rmse_usd_vg,
  mae       = mae_usd_vg
)
```

### Model 5: SARIMA (USD)

```{r plot-sarima-usd}
plot_forecast(
  dates     = sarima_dates,
  actual    = actual_usd_sar,
  predicted = pred_usd_sar,
  title     = "Model 5: SARIMA",
  y_label   = "BTC Price (USD)",
  rmse      = rmse_usd_sar,
  mae       = mae_usd_sar
)
```

### Model 6: BSTS (USD)

```{r plot-bsts-usd}
plot_forecast(
  dates     = bsts_dates,
  actual    = actual_usd_bsts,
  predicted = pred_usd_bsts,
  title     = "Model 6: BSTS",
  y_label   = "BTC Price (USD)",
  rmse      = rmse_usd_bsts,
  mae       = mae_usd_bsts
)
```



## Side-by-Side Panel: All Models (USD)

All four models on the same USD price axis for direct comparison.

```{r panel-usd-var, fig.height=6}
# ── Slide 1: VAR & VAR+GARCH ─────────────────────────────────────────────────
panel_var <- bind_rows(
  data.frame(Date      = var_dates,
             Actual    = actual_usd_var,
             Predicted = pred_usd_var,
             Model     = sprintf("VAR Multi-Step  |  RMSE=$%s  |  MAE=$%s",
                                 format(round(rmse_usd_var), big.mark = ","),
                                 format(round(mae_usd_var),  big.mark = ","))),
  data.frame(Date      = var_dates,
             Actual    = actual_usd_vg,
             Predicted = pred_usd_vg,
             Model     = sprintf("VAR+GARCH Rolling  |  RMSE=$%s  |  MAE=$%s",
                                 format(round(rmse_usd_vg), big.mark = ","),
                                 format(round(mae_usd_vg),  big.mark = ",")))
)
panel_var$Model <- factor(panel_var$Model, levels = unique(panel_var$Model))

panel_var_long <- reshape2::melt(panel_var, id.vars = c("Date", "Model"),
                                 variable.name = "Series", value.name = "USD")

ggplot(panel_var_long, aes(x = Date, y = USD, colour = Series)) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  facet_wrap(~Model, ncol = 2, scales = "fixed") +
  scale_colour_manual(values = c("Actual" = "#2166ac", "Predicted" = "#d6604d")) +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "top",
        strip.text        = element_text(face = "bold", size = 11),
        panel.grid.minor  = element_blank()) +
  labs(title    = "VAR Models: Actual vs. Predicted BTC Price (USD)",
       subtitle = "Test period: June 2025 – January 2026",
       x = "Date", y = "BTC Price (USD)", colour = "")
```

```{r panel-usd-logprice, fig.height=6}
# ── Slide 2: SARIMA & BSTS ───────────────────────────────────────────────────
panel_lp <- bind_rows(
  data.frame(Date      = sarima_dates,
             Actual    = actual_usd_sar,
             Predicted = pred_usd_sar,
             Model     = sprintf("SARIMA  |  RMSE=$%s  |  MAE=$%s",
                                 format(round(rmse_usd_sar), big.mark = ","),
                                 format(round(mae_usd_sar),  big.mark = ","))),
  data.frame(Date      = as.Date(bsts_dates),
             Actual    = actual_usd_bsts,
             Predicted = pred_usd_bsts,
             Model     = sprintf("BSTS  |  RMSE=$%s  |  MAE=$%s",
                                 format(round(rmse_usd_bsts), big.mark = ","),
                                 format(round(mae_usd_bsts),  big.mark = ",")))
)
panel_lp$Model <- factor(panel_lp$Model, levels = unique(panel_lp$Model))

panel_lp_long <- reshape2::melt(panel_lp, id.vars = c("Date", "Model"),
                                variable.name = "Series", value.name = "USD")

ggplot(panel_lp_long, aes(x = Date, y = USD, colour = Series)) +
  geom_line(linewidth = 0.8, alpha = 0.85) +
  facet_wrap(~Model, ncol = 2, scales = "fixed") +
  scale_colour_manual(values = c("Actual" = "#2166ac", "Predicted" = "#d6604d")) +
  scale_y_continuous(labels = scales::dollar) +
  theme_minimal(base_size = 12) +
  theme(legend.position  = "top",
        strip.text        = element_text(face = "bold", size = 11),
        panel.grid.minor  = element_blank()) +
  labs(title    = "Log-Price Models: Actual vs. Predicted BTC Price (USD)",
       subtitle = "Test period: June 2025 – January 2026",
       x = "Date", y = "BTC Price (USD)", colour = "")
```



## Model Comparison Summary

```{r model-comparison}
# USD-scale metrics are directly comparable across all models
results_usd <- data.frame(
  Model = c("VAR (multi-step)", "VAR+GARCH (rolling 1-step)", "SARIMA", "BSTS"),
  Forecast_Scale = c("Returns → USD", "Returns → USD", "log-price → USD", "log-price → USD"),
  RMSE_USD = paste0("$", format(round(c(rmse_usd_var, rmse_usd_vg, rmse_usd_sar, rmse_usd_bsts)), big.mark = ",")),
  MAE_USD  = paste0("$", format(round(c(mae_usd_var,  mae_usd_vg,  mae_usd_sar,  mae_usd_bsts)),  big.mark = ","))
)

print(results_usd)

# Native-scale metrics (for reference / methodological discussion)
cat("\n--- Native-scale metrics (for reference) ---\n")
cat("Returns scale (VAR, VAR+GARCH):\n")
print(data.frame(
  Model = c("VAR (multi-step)", "VAR+GARCH (rolling 1-step)"),
  RMSE  = round(c(rmse_var, rmse_rolling), 4),
  MAE   = round(c(mae_var,  mae_rolling),  4)
))

cat("\nLog-price scale (SARIMA, BSTS):\n")
print(data.frame(
  Model = c("SARIMA", "BSTS"),
  RMSE  = round(c(rmse_sarima, rmse_bsts), 4),
  MAE   = round(c(mae_sarima,  mae_bsts),  4)
))
```

## Key Findings

- **H1 (ARCH effect): Accepted**
The ARCH LM test on raw BTC returns (χ² = 115.36, p < 2.2e-16) confirms strong
volatility clustering. ARCH(1) residuals retained significant ARCH effects
(p = 6.6e-11), confirming it was insufficient. GARCH(1,1) with Student-t errors
eliminated remaining effects (p = 0.856), better capturing heavy tails and
persistent volatility.

- **H2 (Macro linkage): Rejected**
Pairwise Granger tests at α = 0.05 found no evidence that any of the six
macro/market variables (rSPY, rQQQ, dFF, dGS10, dCPI, dUNR) improve forecasts
of BTC returns beyond BTC's own lags (all p > 0.16). This suggests BTC is driven
more by crypto-specific sentiment and liquidity than broader financial conditions,
and limits the benefit of VAR and BSTS models whose strength depends on
informative exogenous variables.

- **H3 (Static VAR vs. Rolling VAR+GARCH): Rejected**
The rolling 1-step-ahead VAR+GARCH (RMSE = 1.9296) marginally underperformed the
static VAR multi-step forecast (RMSE = 1.8726) on the USD scale. The overhead of
re-estimation did not compensate for the lack of predictive signal in the
macro/market variables, consistent with the Granger causality results.

- **H4 (BSTS flexibility): Accepted**
On the log-price scale, BSTS (RMSE = 0.1060, MAE = 0.0914) outperformed SARIMA
(RMSE = 0.2416, MAE = 0.1665) by ~56%. By jointly modeling latent trend,
seasonality, and regression components with Bayesian variable selection, BSTS
adapts to structural changes and captures slower-moving dynamics that a purely
autoregressive SARIMA cannot.


# Future Work

1. **Richer Feature Set:**
The rejection of H2 suggests the current macro variables lack predictive power
over BTC returns. Future work could incorporate crypto-native features — on-chain
metrics (hash rate, active addresses, exchange inflows), BTC options-implied
volatility, and social media sentiment — which may better capture the speculative
dynamics that drive BTC price movements.

2. **Nonlinear and Deep Learning Models:**
All models in this project assume linear dynamics in the mean equation. LSTM or
Transformer-based models (e.g., Temporal Fusion Transformer) can capture
nonlinear dependencies and long-range temporal patterns that VAR and SARIMA
cannot, and are increasingly standard benchmarks in cryptocurrency forecasting.

3. **Regime-Switching Models:**
The 6-month test window (June 2025 – January 2026) spans a single market regime.
A Markov-switching GARCH or threshold VAR could detect and adapt to structural
shifts between bull and bear regimes, which are a defining feature of BTC's price
history and a likely explanation for why static model parameters underperform.

4. **Trading Strategy Backtesting:**
Forecast accuracy metrics (RMSE, MAE) measure statistical performance but not
economic value. Converting model forecasts into a long/short trading rule and
evaluating risk-adjusted returns — Sharpe ratio, maximum drawdown, win rate —
over the test period would directly address the original objective of identifying
profitable trading opportunities.
